Language-Parametric Techniques for Live Probes

### 2023-07-25 paper/paper.tex
add steps

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=engineering,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Live programming, language engineering},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

@@

  % What argument, feasibility proof, artifacts, or results and evaluation support this work?
  \paragraph{Grounding}
  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.  We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead. Our results show that live probes on top of native debugger APIs are performant enough for interactive use. In the case of DAP, it depends on characteristics of the programming language.

@@

The question left unanswered in Victor's presentation is: how to build such live probes? While there has been recent research showcasing implementation approaches~\cite{ExampleBasedGraal}, they depends on a specific runtime system (e.g., GraalVM), or language specific ``hacks''~\cite{LiveLit}, and hence cannot be easily transferred to other languages. So a more specific question is: how to  engineer live probes \textit{generically}, without dedicated compiler or runtime support?

@@

\section{Background and Overview}
\label{sec:problem-overview}

The basic concept of live probes is illustrated in Figure~\ref{FIG:bret}, showing two panes. 
The left pane shows an editor containing the source code of a possible implementation of binary search in Javascript. 
The right pane shows the values of variables matched up to the source line they were last updated.
The top two rows specify example input data that the function is executed on. The consecutive values of variables that are updated in a loop (e.g. \lstinline{low}, \lstinline{high}, \lstinline{mid} and \lstinline{value}) are displayed adjacently. So for \lstinline{low}, the history of values is 0, 3, and 5. Note that the update of \lstinline{high} in the else-if branch (apparently) never executed, so there is no (redundant) value display in the right pane. Finally, the last line shows the final return value of the function: apparently the \lstinline{key} was not found in the \lstinline[language=c]{array}.

What makes this \textit{live} is that while the programmer is editing the program on the left (or the input arguments to the function), the value display on the right continuously updates so that it is apparent what is going on at run time.

Dissecting this example leads to the following technical requirements for implementing live probes:
\begin{itemize}
  \item there should a way to specify the inputs to a method
  \item after every change to the code, it needs to be recompiled and reexecuted
  \item during execution the values of variables needs to be recorded and linked to source lines
  \item the resulting history list of values needs to be communicated to the IDE
\end{itemize}

The basic intuition of our approach to meet these requirements is as follows: a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code. Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger. It then queries the source code for annotations (either explicit annotations as in Java, or special comments) specifying the input arguments to the method. The KAA then sets a breakpoint on the first statement of the method, and invokes it. Since there is a breakpoint, the debugger halts. Using the common step-over facility of a typical debugger, the KAA then steps through the whole method until its execution completes. During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables. These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLit,ExampleBasedGraal}, or using some other UI affordance.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{img/bret}
  \caption{Screenshot of Bret Victor's 2012 talk ``Inventing on Principle'' (23rd minute)}
  \label{FIG:bret}
\end{figure}

@@

To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
The scenario describes the programming of a function performing a binary search on an array of characters. It consists of 18 steps, including 10 steps where the source code is modified and 8 steps where the parameters are changed. 
The scenario also includes the case where the tested method does not finish. As the code in Bret Victor's presentation is in Javascript, it has been adapted to the 3 languages studied in this paper.

@@

LiveLiterals (Van der Storm, Live 2016)

>>>>

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Social Coding, General-purpose programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

@@

  % What argument, feasibility proof, artifacts, or results and evaluation support this work?
  \paragraph{Grounding}
  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, and C, all with minimum glue code.  We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead. Our results show that live probes on top of native debugger APIs are performant enough for interactive use. In the case of DAP, it depends on characteristics of the programming language.

@@

The question left unanswered in Victor's presentation is: how to build such live probes? While there has been recent research showcasing an implementation approach~\cite{ExampleBasedGraal}, it depends on specific a runtime system (GraalVM), and cannot be easily transferred to other languages. So a more specific question is: how to  engineer live probes without dedicated compiler or runtime support?

@@

\section{Problem Overview}
\label{sec:problem-overview}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{img/bret}
  \caption{Screenshot of Bret Victor's 2012 talk ``Inventing on Principle'' (23rd minute)}
  \label{FIG:bret}
\end{figure}

@@

To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
The scenario describes the programming of a function performing a binary search on an array of characters. It consists of 19 steps, including 10 steps where the source code is modified and 9 steps where the parameters are changed.

\begin{enumerate}
  \item The function is defined and the parameters are defined to be an array of characters(\code{['a','b','c','d','e',f']}) and a target character(\code{'d'}).
  \item A new variable \code{low} is defined.
  \item A new variable \code{high} is defined.
  \item A new variable \code{mid} is defined.
  \item The variable \code{mid} is changed to be a integer.
  \item A new variable \code{value} is defined.
  \item A if and else snippets is added.
  \item The target is changed to \code{'b'}.
  \item The target is changed to \code{'c'}.
  \item The target is changed to \code{'d'} and the code from the \code{mid} is refactored to be in a \code{while(true)} loop.
  \item The target is changed to \code{'a'}.
  \item The target is changed to \code{'b'}.
  \item The target is changed to \code{'c'}.
  \item The target is changed to \code{'d'}.
  \item The target is changed to \code{'e'}.
  \item The target is changed to \code{'f'}.
  \item The target is changed to \code{'g'}.
  \item The condition of the while loop is changed to \code{low <= high}.
  \item A return \code{-1} is added at the end of the function.
\end{enumerate}

The scenario also includes the case where the tested method does not finish at step 17. As the code in Bret Victor's presentation is in Javascript, it has been adapted to the 3 languages studied in this paper.

@@
